From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-27 upi.tex
done with sec 2 for now

\newcommand{\byiso}[1]{{\leftrightarrow}{\langle} ~#1~ \rangle}
\newcommand{\byisotwo}[1]{{\Leftrightarrow}{\langle} ~#1~ \rangle}
\newcommand{\unitepl}{\texttt{unitepl}}
\newcommand{\unitipl}{\texttt{unitipl}}
\newcommand{\unitepr}{\texttt{unitepr}}
\newcommand{\unitipr}{\texttt{unitipr}}
\newcommand{\swap}{\texttt{swap}}
\newcommand{\swapp}{\texttt{swapp}}
\newcommand{\assoclp}{\texttt{assoclp}}
\newcommand{\assocrp}{\texttt{assocrp}}
\newcommand{\unitetl}{\texttt{unitetl}}
\newcommand{\unititl}{\texttt{unititl}}
\newcommand{\unitetr}{\texttt{unitetr}}
\newcommand{\unititr}{\texttt{unititr}}
\newcommand{\swapt}{\texttt{swapt}}
\newcommand{\assoclt}{\texttt{assoclt}}
\newcommand{\assocrt}{\texttt{assocrt}}
\newcommand{\absorbr}{\texttt{absorbr}}
\newcommand{\absorbl}{\texttt{absorbl}}
\newcommand{\factorzr}{\texttt{factorzr}}
\newcommand{\factorzl}{\texttt{factorzl}}
\newcommand{\factor}{\texttt{factor}}
\newcommand{\distl}{\texttt{distl}}
\newcommand{\dist}{\texttt{dist}}
\newcommand{\factorl}{\texttt{factorl}}
\newcommand{\id}{\texttt{id}}
\newcommand{\compc}[2]{#1 \circ #2}
\newcommand{\compcc}[2]{#1 \bullet #2}
\newcommand{\respcomp}[2]{#1 \odot #2}

@@

Having illustrated the general flavor of the $\Pi$ family of
languages, we present in full detail a small $\Pi$-based language
which we will use in the formalization in the rest of the paper. The
language below is the restriction of $\Pi$ to the case of just one
type, the type of booleans.

\[\def\arraystretch{0.8}\begin{array}{rcl}
\tau &::=& \bt \\
\\
v &::=& \begin{array}[t]{lrcl}
                    & \fc &:& \bt \\
              \alt & \tc &:& \bt
               \end{array} \\
\\
c &::=& \begin{array}[t]{llcl}
              & \id &:& \tau \iso \tau \\
               \alt & \swap &:& \bt \iso \bt \\
               \alt & ! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
               \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3) \to (\tau_1 \iso \tau_3)
               \end{array} \\
\\
\alpha &::=& \begin{array}[t]{lrcl}
               & \id &:& c \isotwo c \\
            \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
            \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
            \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
            \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
            \alt & \rho &:& \swap \circ \swap \isotwo \id \\
            \alt & \assocc &:& \compc{(\compc{c_1}{c_2})}{c_3} \isotwo \compc{c_1}{(\compc{c_2}{c_3})} \\
            \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2') \to (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
            \alt & !! &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_1) \\
            \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3) \to (c_1 \isotwo c_3)
             \end{array}
\end{array}\]

Canonical forms: given a level 1 program p, the level 2 combinators are
complete; they can transform p to either id or not:
\begin{verbatim}
cmpl2-lem : (p : B <->1 B) -> (p <->2 id) + (p <->2 not)
\end{verbatim}

>>>>

\newcommand{\byiso}[1]{{\leftrightarrow}{\langle} ~#1~ \rangle}
\newcommand{\byisotwo}[1]{{\Leftrightarrow}{\langle} ~#1~ \rangle}
\newcommand{\unitepl}{\texttt{unitepl}}
\newcommand{\unitipl}{\texttt{unitipl}}
\newcommand{\unitepr}{\texttt{unitepr}}
\newcommand{\unitipr}{\texttt{unitipr}}
\newcommand{\swap}{\textit{swap}}
\newcommand{\swapp}{\texttt{swapp}}
\newcommand{\assoclp}{\texttt{assoclp}}
\newcommand{\assocrp}{\texttt{assocrp}}
\newcommand{\unitetl}{\texttt{unitetl}}
\newcommand{\unititl}{\texttt{unititl}}
\newcommand{\unitetr}{\texttt{unitetr}}
\newcommand{\unititr}{\texttt{unititr}}
\newcommand{\swapt}{\texttt{swapt}}
\newcommand{\assoclt}{\texttt{assoclt}}
\newcommand{\assocrt}{\texttt{assocrt}}
\newcommand{\absorbr}{\texttt{absorbr}}
\newcommand{\absorbl}{\texttt{absorbl}}
\newcommand{\factorzr}{\texttt{factorzr}}
\newcommand{\factorzl}{\texttt{factorzl}}
\newcommand{\factor}{\texttt{factor}}
\newcommand{\distl}{\texttt{distl}}
\newcommand{\dist}{\texttt{dist}}
\newcommand{\factorl}{\texttt{factorl}}
\newcommand{\id}{\textit{id}}
\newcommand{\compc}[2]{#1 \circ #2}
\newcommand{\compcc}[2]{#1 \bullet #2}
\newcommand{\respcomp}[2]{#1 \odot #2}

@@

Having illustrated the general flavor of the $\Pi$ family of
languages, we present in full detail a small $\Pi$-based language
which we will use in the formalization in the rest of the paper. The
language below is the restriction of $\Pi$ to the case of just one
type, the type of booleans:

\[\def\arraystretch{0.8}\begin{array}{l@{\quad}rclrl}
(\textit{Types}) & \tau &::=& \bt \\
\\
(\textit{Terms}) &  v &::=& \fc &:& \bt \\
              && \alt & \tc &:& \bt \\
\\
 (\textit{1-combinators}) &  c &::=& \id &:& \tau \iso \tau \\
               && \alt & \swap &:& \bt \iso \bt \\
               && \alt & ! &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_1) \\
               && \alt & \circ &:& (\tau_1 \iso \tau_2) \to (\tau_2 \iso \tau_3) \to (\tau_1 \iso \tau_3)  \\
\\
(\textit{2-combinators}) & \alpha &::=& \id &:& c \isotwo c \\
            && \alt & \idlc &:& \compc{\id}{c} \isotwo c \\
            && \alt & \idrc &:& \compc{c}{\id} \isotwo c \\
            && \alt & \invl &:& \compc{c\;}{\;\invc{c}} \isotwo \id \\
            && \alt & \invr &:& \compc{\invc{c}}{c} \isotwo \id \\
            && \alt & \rho &:& \swap \circ \swap \isotwo \id \\
            && \alt & \assocc &:& 
                                  \compc{(\compc{c_1}{c_2})}{c_3} \isotwo \compc{c_1}{(\compc{c_2}{c_3})} \\
            && \alt & \odot &:& (c_1 \isotwo c_1') \to (c_2 \isotwo c_2') \to 
                             (\compc{c_1}{c_2} \isotwo \compc{c_1'}{c_2'}) \\
            && \alt & !! &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_1) \\
            && \alt & \bullet &:& (c_1 \isotwo c_2) \to (c_2 \isotwo c_3) \to (c_1 \isotwo c_3)
\end{array}\]

The syntactic category $c$ is that of 1-combinators denoting
reversible programs, type isomorphisms, permutations, or equivalences
depending on one's favorite interpretation. There are two primitive
combinators $\id$ and $\swap$ which are closed under inverses $!$ and
sequential composition $\circ$. The syntactic category $\alpha$ is
that of 2-combinators denoting reversible program transfomations,
coherence conditions on type isomorphisms, equivalences between
permutations, or program optimizations depending on one's favorite
interpretation.

It is relatively easy to think of a few sound program transformations
to include in the category $\alpha$ of 2-combinators. The set above
however is \emph{complete}.

\begin{lemma}[Canonical Forms]
  Given a 1-combinator $c : \tau \leftrightarrow \tau$, we either have
  a normalization $c \Leftrightarrow \id$ or a normalization
  $c \Leftrightarrow \swap$.
\end{lemma}
